---
title: "Class 09: Halloween Mini-Project"
author: "Eli Haddad (A16308227)"
format: pdf
---

# 1. Importing candy data


```{r}
candy_file <- "candy-data.csv"

candy = read.csv(candy_file, row.names=1)
head(candy)


```
>Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

# 2. What is your favorite candy?

>Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite candy is Milky Way. It's winpercent value is:
```{r}
candy["Milky Way",]$winpercent
```

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```
>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

```{r}
library("skimr")
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent is the one that is different on scale. It is the only column whose max value is not 1.


> Q7. What do you think a zero and one represent for the candy$chocolate column?

I think that the 0 means false and the 1 means true in the candy$chocolate column. For example, if a candy identifies as chocolate, it will have a 1 in this column.

> Q8. Plot a histogram of winpercent values

```{r}
hist(candy$winpercent)
```
> Q9. Is the distribution of winpercent values symmetrical?

The distribution is not symmetrical. It is slightly skewed right. 

> Q10. Is the center of the distribution above or below 50%?

The center of the distribution is below 50%.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}

mean(candy$winpercent[as.logical(candy$chocolate)])
mean(candy$winpercent[as.logical(candy$fruity)])


```
On average, chocolate candy is ranked higher than fruity candy.


> Q12. Is this difference statistically significant?

```{r}
t.test(candy$winpercent[as.logical(candy$chocolate)],candy$winpercent[as.logical(candy$fruity)])
```

The difference is statistically different.

# 3. Overall Candy Rankings

> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy %>% arrange(winpercent) %>% head(5)
```
The five least liked candy types are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.
I prefer the dplyr way because it is a lot more readable and clean.

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>% arrange(-winpercent) %>% head(5)
```
Top 5 all time favorite candy types are Reese's Peanut Butter cup, Reese's Miniatures, Twix, Kit Kat, and Snickers

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
library(ggplot2)


ggplot(candy) + 
  aes(winpercent, rownames(candy)) +
  geom_col()

```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()

```

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```
> Q17. What is the worst ranked chocolate candy?

The worst ranked chocolate candy is Sixlets.

> Q18. What is the best ranked fruity candy?

The best ranked fruity candy is Starburst.

# 4. Taking a look at pricepercent

```{r}
library(ggrepel)

ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```
> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's Miniatures is highest ranked in terms of winpercent for the least money.

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
candy %>% arrange(-pricepercent) %>% head(5)
```
The top 5 are Nik L Nip, Nestle Smarties, Ring pop, Hershey's Krackel, and Hershey's Milk Chocolate. The least popular of these 5 are Nik L Nip.

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment()

```{r}
ggplot(candy) + 
  aes(pricepercent, rownames(candy)) +
  geom_col()
```
```{r}
ggplot(candy) + 
  aes(pricepercent, reorder(rownames(candy),pricepercent)) +
  geom_col()

```
```{r}
ggplot(candy) +
  aes(pricepercent, reorder(rownames(candy), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()

```

# 5. Exploring the correlation structure

```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)

```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The fruity and chocolate variables are most anti-correlated.

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent are most positively correlated, suggesting that chocolate is a favorite.

# 6. Principal Component Analysis

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```
```{r}
plot(pca$x[,"PC1"], pca$x[,"PC2"], xlab="PC1", ylab= "PC2")
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p


```
```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538 Halloween Candy Dataset")
```

```{r}
library(plotly)
#ggplotly(p)

```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The variables strongly picked up by PC1 in the positive direction are fruity, hard, and pluribus. This makes sense to me because it describes that fruity candy are hard and come in a bag of multiple candies. In addition, chocolate and bar are in the opposite direction, which makes sense because chocolate and fruity were negatively correlated with one another in a past plot and it makes sense that chocolate is made in bars.