---
title: "Class 11 Part 2 Population analysis"
author: "Eli Haddad (A16308227)"
format: pdf
---

One sample is obviously not enough to know what is happening in a population. You are
interested in assessing genetic differences on a population scale. So, you processed about
~230 samples and did the normalization on a genome level. Now, you want to find whether
there is any association of the 4 asthma-associated SNPs (rs8067378...) on ORMDL3
expression.
https://bioboot.github.io/bggn213_W19/class- material/rs8067378_ENSG00000172057.6.txt
This is the final file you got (
column is genotype and the third column are the expression values.
). The first column is sample name, the second

> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes

Importing the data
```{r}
data <- read.table("rs8067378_ENSG00000172057.6.txt", sep = " ")
head(data)

```
The sample size for each genotype

```{r}
table(data$geno)
```

Corresponding median expression levels for each of these genotypes

```{r}
boxinfo <- boxplot(data[data$geno == "A/A", ]$exp)
medium_val <- boxinfo$stats[3]

medium_val

```

A/A: 31.24847

```{r}
boxinfo <- boxplot(data[data$geno == "A/G", ]$exp)
medium_val <- boxinfo$stats[3]

medium_val
```


A/G: 25.06486


```{r}
boxinfo <- boxplot(data[data$geno == "G/G", ]$exp)
medium_val <- boxinfo$stats[3]

medium_val
```


G/G: 20.07363


> Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3? 

```{r}
library(ggplot2)

ggplot(data, aes(geno,exp)) +
  geom_boxplot(notch=TRUE, aes(fill=geno)) + 
  geom_jitter(width = 0.2, alpha=0.5) +
  labs(title = "Expression Data of Various SNPs", x = "Genotype", y = "Expression")


```

From the boxplot, I can infer that A/A has relatively greater expression value than G/G, as the median expression of A/A is greater than that of G/G. Since the notches between the boxplots of A/A and G/G do not overlap, this difference is statistically different, suggesting that the SNP does effect the expression of ORMDL3.


